---
title: "Summarizing categorical data "
subtitle: "STAT 311 Spring 2024"
author: "YOUR NAME HERE"
date: "`r Sys.Date()`"
output:    
      html_document:
         toc: yes
         toc_float: yes
---

## Introduction

Today we will look at traffic accident data in Seattle.
The data are open and made available in a comma-separated-value (CSV) format by [Seattle Geodata](https://data-seattlecitygis.opendata.arcgis.com/datasets/5b5c745e0f1f48e7a53acec63a0022ab_0). It covers all recorded accidents in Seattle from Jan 1 2020 through Dec 16 2020; some of the variables were deleted for the purposes of this exercise.

The goal for today is conduct a preliminary exploratory analysis of these data, with particular emphasis on summarizing a few categorical variables graphically and also in a tabular format.

<!--- 

The packages used for this analysis are loaded in the code chunk shown below. We will also need to read in the dataset `accidents.csv` since it is not part of any package. We can read it using the `read_csv` function from the **readr** package which is part of the **tidyverse** ecosystem. For more information on importing datasets in R, please see [chapter 7](https://r4ds.hadley.nz/data-import) in the excellent open source reference **R for Data Science**.  

--->


```{r setup, echo = FALSE, message = FALSE, warning = FALSE}
#load packages and set global options
library(tidyverse)        #umbrella package
library(janitor)          #for tables 

accidents <- read_csv("accidents.csv")
```

Below is an excerpt of the data dictionary also known as the code book.  Note that it is long (there are lots of variables in the data), but we will be using a limited set of the variables in this tutorial.

| Header                        |  Description
|:----------------|:--------------------------------
| `id` | unique identifier
| `addr_type` | Collision address type: **Alley**, **Block**, **Intersection** 
| `location` | Text description of the general location of the collision
| `severity_code` |3 - fatality, 2b - serious injury, 2 -injury, 1 - property damage, 0 - unknown
| `severity_desc   ` | Description of severity: **Fatality Collision**, **Injury Collision**, **Property Damage Only Collision**, **Serious Injury Collision**, **Unknown**
| `collision_type` | Collision type: **Angles**, **Cycles**, **Head On**, **Left Turn**, **Parked Car**, **Pedestrian**, **Rear Ended**, **Right Turn**, **Sideswipe**
| `person_count` | Number of people involved in the collision
|`ped_count` |Number of pedestrians involved in the collision
| `pedcyl_count` | Number of bicycles involved in the collision 
|`veh_count`| Number of vehicles involved in the collision (0, 1, 2, etc)
|`injuries`|Number of total injuries in the collision
|`serious_injuries`|Number of serious injuries in the collision
|`fatalities`| Number of fatalities in the collision
|`inc_date`| Date of incident
|`under_infl`| Whether or not the driver was under the influence of drugs or alcohol: **Y**, **N**
|`weather`|A text description of the weather conditions during the collision
|`road`|The condition of the road during the collision: **Dry**, **Wet**, **Ice**, **Sand/Mud/Dirt**, **Snow/Slush**, **Standing Water**, **Unknown**
|`light`|Light conditions during the collision: **Dark - No Street Lights**, **Dark - Street Lights Off**, **Dark - Street Lights On**, **Dark - Unknown Lighting**, **Dusk**, **Dawn**, **Daylight**
|`speeding`|  Whether or not speeding was a factor in the collision: **Y**, **N**
|`hit_parked_car`|Whether or not the collision involved hitting a parked car: **Y**, **N**


The following provides a snapshot of the variable types in the data.

```{r peek-at-data, echo = FALSE, message = FALSE, warning = FALSE}
glimpse(accidents)

```


* * * 

## Warm Up

Before proceeding, load the packages and read in the data to  your working environment by running the first code chunk labeled as `setup`. Then open the dataset in the spreadsheet viewer to take a look.

Let's do some warm up exercises to familiarize ourselves with the data.

1. What does each row represent? A recorded accident, the registration number of a car, the location of an accident, an insurance claim?


2. The variable `veh_count` describes the number of cars involved in the collision. How many accidents involved 2 or more cars?


3. The variable `addr_type` describes the location of the accident. Where do most accidents occur?  Create a frequency table of `addr_type` in descending order of frequency.


* * * 

## Data Analysis


Let's now examine the relationship between the light conditions (`light`) and whether or not it is related to parked cars being hit (`hit_parked_car`). 

1.  We begin by creating a frequency table of each variable to examine their distributions. Go ahead and write code to do this.

    ```{r 1a}

    accidents %>% count(light)

    accidents %>% count(hit_parked_car)

    ``` 

2. As we can see, the `light` variable has three categories of darkness, two of which are quite sparse. Perhaps we can combine all three into one category and call it "Dark"?
 
    Fill in the blanks in the code chunk to create and add a new variable, called `light_cat`, which is the same as `light` except that it combines all the "Dark" categories together into one category called "Dark". Then examine the frequency distribution of `light_cat`.


    ```{r recode-light}

     accidents <- accidents %>% 
             mutate( light_cat = 
                       ifelse(
                           light == "Dark - Street Lights Off" | 
                           light == "Dark - Street Lights On" | 
                           light == "Dark - No Street Lights", "Dark", light) )

    ```


3. Now let us examine whether light conditions influence whether or not a parked car is hit. Create a contingency table of the frequencies for `light_cat` and `hit_parked_car`. If there is no association between these variables, what % of cars should be hit regardless of light conditions?


    ```{r }

    accidents %>% 
   tabyl( light_cat, hit_parked_car, show_na = FALSE) %>% 
   adorn_totals(where = "row") %>%
   adorn_percentages() %>%
   adorn_pct_formatting(digits=2) %>%
   adorn_ns() %>%
   adorn_title()
    ```


     If there is no association between the light conditions and whether or not a parked car is hit, we should see 11.88% of cars being hit in each row. Clearly, daylight hours have fewer such incidents and the hours when there is no/less light have more such incidents. There is an association between these variables.


4. Make a segmented bar plot to visualize the contingency table from part 3 by filling in the blanks in the code below.


    ```{r seg-bar-light-hit}

    
    ggplot( data = accidents , 
        mapping = aes( x = light_cat, 
                       fill = hit_parked_car ) ) +
     geom_bar( position = "fill" ) +
    labs(x= "light condition", 
         y = "proportion",
         fill = "hit parked car", 
        title = "Light conditions vs. hitting parked cars")

    
    
    ```


5. Make a mosaicplot to visualize the strength of the associations seen in the contingency table. Write a couple of sentences to describe what you learn.

    ```{r}

mosaicplot(hit_parked_car ~ light_cat, data = accidents, shade = TRUE)

    ```

    We knew from the contingency table that the dark hours have a higher rate of parked cars being hit than the daylight hours. The mosaic plot confirms that the numbers in these cells are statistically meaningful. 


## More practice

Do as many as you feel comfortable with.

Suppose we want to know if `addr_type` is associated with `severity_desc`. Read about these variables in the code book. Then do the following.

6. Make a frequency distribution of each variable separately.

    ```{r}
    accidents %>% count(addr_type)

    accidents %>% count(severity_desc)

    ```


7. Create and add a new variable to the `accidents` dataframe called `severity` variable which takes the same values as `severity_desc`, except that it combines the categories "Injury Collision" and "Serious Injury Collision" into one category called "Injury collision".

    ```{r} 
    accidents <- accidents %>% 
             mutate(severity = 
              ifelse(severity_desc == "Injury Collision" |
                      severity_desc == "Serious Injury Collision", 
                                        "Injury collision",
                                         severity_desc) )
    
    ```


8. Create a contingency table of the frequencies for `addr_type` and `severity_desc`. If there is no association between these variables, what % of collisions should be injury collisions regardless of address type?

    *Hint* think about whether you want `addr_type` on the rows or the column. 
    
    ```{r}
   accidents %>% 
   tabyl( addr_type, severity, show_na = FALSE) %>% 
   adorn_totals(where = "row") %>%
   adorn_percentages() %>%
   adorn_pct_formatting(digits=2) %>%
   adorn_ns() %>%
   adorn_title()

    ```


9. Make a segmented bar plot to visualize the contingency table from part 8.

    ```{r}
    ggplot( data = accidents %>% filter(!is.na(addr_type)), 
        mapping = aes( x = addr_type, 
                       fill = severity) ) +
     geom_bar( position = "fill" ) +
    labs(x= "address type", 
         y = "proportion",
         fill = "severity", 
        title = "Severity of collision vs. address type")

    
    ```


10. Make a mosaicplot to visualize the contingency table from part 8.

    ```{r}
    mosaicplot(severity ~ addr_type, data = accidents, shade = TRUE)

    ```