---
title: "Bootstrapping the Class Survey "
subtitle: "Ranjini Grove"
date: "`r Sys.Date()`"
output:    
      html_document:
         toc: no
         toc_float: no
      editor_options:
         chunk_output_type: console
---

In this report we will analyze survey responses from STAT 311 students collected at the start of spring quarter 2024. Specifically, we are interested in going beyond summaries to make inferences about all UW students. We will primarily focus on constructing interval estimates for population  values using the bootstrap methods.



The packages we will be using have been loaded in the code chunk below. You can add additional packages as necessary. 

```{r load-packages, include=FALSE}
library(tidyverse)      #umbrella suite
library(tidymodels)     #for bootstrapping with infer
library(skimr)          #for skimming variables

```

## Data

The class data was collected via a Google form and was downloaded as a CSV file named `classdata_spr2024.csv`. The codebook with a description of the variables in the dataset is shown below. The overall data cleaning steps were extremely minimal. I renamed the variables so they were shorter. Additional data cleaning and recoding is done for the variables involved in specific analyses and these will be described in the relevant sections. All the code is included in the Rmarkdown document. 

```{r data-cleaning, include = FALSE}
classdata <- read_csv(file="classdatasp24.csv")

classdata <- classdata %>%
             rename( class = contains("class standing"),
               credits = contains("credits"),
               diet = contains("vegetarian"),
               sleep = contains("sleep"),
               dating_pref=contains("attractive"),
               sex=contains("sex"),
               gender=contains("gender"),
               study=contains("study"),
               birthday=contains("born"),
               section=contains("quiz section"),
               haircut=contains("haircut"),
               shoes=contains("shoes"),
               nervous=contains("nervous"),
               siblings=contains("siblings"),
               which_tire=contains("tire"),
               height=contains("height"),
               speed=contains("fastest"),
               years_in_wa=contains("WA state"),
               snapchat=contains("snapchat"),
               insta=contains("instagram"),
               travel=contains("visited"),
               salary=contains("earn"),
               breakfast=contains("breakfast"),
               eyewear=contains("eyeglasses"),
               rent=contains("rent"),
               climate_worry = contains("global warming") ) 

classdata <- classdata %>% 
           mutate(dating_pref = na_if(dating_pref, "yes;No"),
                  sex = na_if(sex, "Prefer not to answer"))

```


|Variable |  Description |
| :----:  |  :--------  |
|Timestamp| Self explanatory |
|class    | What is your class standing?|
|credits  | How many credits are you signed up for in Spr 2024?|
|vegetarian| Are you a vegetarian?|
|sleep    | How many hours of sleep do you get on average at night?|
|dating_pref| Would you date someone with a great personality even if you did not find them attractive? |
| sex     | What is your biological sex? |
|gender   | To which gender do you most closely identify? |
|study    |How many hours do you study outside of classes (on average) per week? (if this is your first quarter at UW, you may skip this question) |
|birthday | What day of the week were you born? |
|section  | Which quiz section are you in? |
|haircut | What is the most (in dollars) you have paid for a haircut? |
|shoes   | How many pairs of shoes do you own? |
|nervous | How nervous are you about taking STAT 311?|
|siblings | How many siblings do you have? (include step sibs too)|
|which_tire| Suppose you were late to an exam and decided to use the "flat tire" excuse. Which tire would you  pick?|
|height | What is your height (in inches)?|
|years_in_wa | How many years have you lived in WA state? (enter 0 if none, round your answer to a whole number)|
|snapchat | How many friends do you have on snapchat? (enter 0 if you don't use it)|
|insta| How many friends do you have on instagram? (enter 0 if you don't use it)|
|travel |How many states in the U.S. have you visited?|
|salary| How much money did you earn last year (in dollars )? |
|breakfast| Do you usually eat breakfast in the morning?|
|eyewear | Do you wear eyeglasses or contact lenses?|
|rent | What is your monthly rent?|
|climate_worry| Scientists predict that global warming may have big effects on the polar regions within the next 100 years. One of the possible effects is that the northern ice cap may completely melt. Would this bother you a great deal, some, a little, or not at all if it actually happened?|




## Exercises

### Part 1: How many credits?

Let's consider responses to this question "How many credits are you signed up for in Spring 2024". 

The first step in any data analysis is to become acquainted with the data. We begin by skimming this variable to get a bird's eye view of the values, missingness and other possible issues. 



1. Make a plot to display the distribution of `credits`. Comment on the shape. 


2. The output from `skim` shows the mean and standard deviation for `credits`. Are they appropriate measures of center and spread for the distribution in exercise 1? Why or why not?

    Regardless of your answer to the previous question, we want to estimate the average (mean) number of credits taken by all students at UW in spring 2024. Our sample estimates this number as `r round( mean(classdata$credits), 2)`. However, estimates from samples can vary (depending on the individuals in our sample), so it is important to assess how this number might vary across samples. 
    
    How can we study sampling variability using our single sample? We will do so using a technique known as the bootstrap method.
    
    The bootstrapping scheme proceeds as follows:

-   **Step 1.** Take a bootstrap sample - a random sample taken **with replacement** from the original sample, of the same size as the original sample.
-   **Step 2.** Calculate the summary - such as mean, median, proportion, slope, etc. computed on the bootstrap samples.
-   **Step 3.** Repeat steps (1) and (2) many times to create a bootstrap distribution - a distribution of values from the resamples.


    Instead of coding up each of these steps, we will create our resamples using the **specify, generate, calculate** workflow from the  **tidymodels** package.


    
3. Fill in the blanks in the code chunk below to implement the bootstrap method to create 1,000 resamples of size $n = 121$ (same size as our sample) and calculate the mean number of `credits` for each resample.  

    ```{r bootstrap-mean-credits, eval = F}
    set.seed (45)            #for reproducibility
    
    boot_mean.df <- ___ %>% 
      specify( response = ___ ) %>% 
      generate(reps = 1000, type = "___") %>% 
      calculate(stat = "___")
    ```

4. Make a histogram to display the distribution of the mean `credits` from the 1,000 resamples. Comment on the shape. 

 

    The histogram of the sample means in exercise 4 is called the **bootstrapped distribution**. It is an approximation to what we would see if we could sample repeatedly from the actual population in the sense that both distributions would have the same shape and spread.
    
    Now, let us return to our goal of estimating the average (mean) number of credits taken by all UW students in spring 2024. Think of this value as the **fish**. On the one hand, we could use our sample mean `r round(mean(classdata$credits), 2)` to estimate this number. This is like fishing with a **spear**. 
    
    What would fishing with a net look like? Look at the bootstrapped distribution from exercise 4. Between which two numbers would you say most of the sample means lie? Think of this interval as the **net**. 
    
    We will now introduce the percentile method for constructing this interval more precisely. It requires specifying a level of confidence that we are aiming for.

5. One method to construct a confidence interval is to use the middle 95% of values of the bootstrap distribution. We can do this by computing the 2.5th and 97.5th percentiles. This is known as the **95% bootstrap percentile interval**. Calculate this interval for the mean number of credits taken by all students at UW.

  

    It turns out that according to UW Academic Data Management, the average number of credits that UW students signed up for in Spring 2024 is 14.23. Does our interval contain this value?
    
    In order to understand the meaning of 95% confidence, re-run the code chunks in exercise 3- 5 several times after changing the seed each time. Does your confidence interval stay the same?
Does it contain the true population value? 

    This is where the 95% confidence comes in. For every 100 intervals that we generate in this manner, we can expect that only about 95% of the intervals we have created will contain the true population value. 
    
    And this is only if we can think of our original sample as truly representing a random subset of all UW students.
    
* * *

**Your Turn**

Calculate a 95% bootstrap percentile interval for the average number of hours of sleep that UW students get per night. Repeat exercises 1-5 for the variable `sleep`.


* * *

The bootstrap method can be used to generate confidence intervals for any population summary, such the mean, median, difference of means, slope etc. 

In the next part we will use the method to bootstrap an interval for a difference of means.

* * *

### Part II: Do men drive faster?

In this section we will compare the responses to the question "What is the fastest you have ever driven?" between males and females.

1. Make a preliminary visualization to explore the relationship between the relevant variables `speed` and `sex`. You can also use `skim`.


2. Calculate the mean speeds reported by male vs. female drivers. Calculate the difference in mean speeds (male - female). 


3. Fill in the blanks in the code chunk below to implement the bootstrap method to create 1,000 resamples and calculate the difference in mean `speed` for each resample. 



    ```{r bootstrap-diff-means, eval = F}
    set.seed(11)

    boot_diff_means.df <- ___ %>% 
                     filter(!is.na(speed), !is.na(sex))    %>%
                       specify( response = ___, explanatory = ___) %>% 
                       generate(reps = ___, type = "___") %>% 
                       calculate(stat = "diff in means", 
                                order = c("Male", "Female"))
   

```

4. Make a histogram to display the distribution of the difference in mean `speed` between males and females from the 1,000 resamples. Comment on the shape. 

    
5. Calculate the 95% bootstrap percentile interval for the difference in mean speeds driven by men vs. women in the population. Is "0" in the interval? What does this tell you?


    
    
**Stretch** Come up with a summary you want to calculate and generate its bootstrapped distribution. 
    