---
title: "A grammar of data wrangling "
subtitle: "STAT 311 Spring 2024"
author: "YOUR NAME HERE"
date: "`r Sys.Date()`"
output:    
      html_document:
         toc: yes
         toc_float: yes
---

Throughout this lesson and the next, we will use `gapminder` dataset which is available in an R package of the same name. The `gapminder` dataframe tracks demographic data in countries of the world over time. You can visit their [website](https://www.gapminder.org) to see the breadth of projects they work on. 

Let's load the packages we will need today. Note that we only need the `dplyr` package for the data manipulation verbs we will be using, however we will usually prefer to load `tidyverse`, the umbrella package which contains `dplyr` within it.

```{r setup, include = FALSE}
#load packages and set global options
library(tidyverse)        #umbrella package
library(gapminder)        #dataset is in this package

```

## Warm Up

To learn more about the data set, you can bring up the help file with `?gapminder`. Let's also load the data using `data` and take a look at it in the spreadsheet editor.

Take a look at the variable types  with the `glimpse` function. 

* * * 

## Part 1: select, filter, mutate, arrange

In this part, we will practice using the data verbs `filter`, `select`, `arrange` and `mutate` in order to answer various queries. 
Fill in the blanks in the code provided in each code chunk to accomplish a specific task. Consult the slidedeck "Introduction to Data: Data Manipulation" for help. Each code chunk has the `eval` option set to FALSE. Once you have filled in the blanks you may delete this option so your code will run when the document is knitted.

1. Use `filter` to extract the data for Canada and select the variables country, continent, year, and lifeExp. Replace [EQUAL TO] with the logical operator for "exactly equal to"


    ```{r part1-ex1, eval = FALSE}
    
    gapminder %>% 
        filter( country [EQUAL TO] "___") %>%
              select( ___, ___, ____, ____ )

    ```

2. Which countries had a life expectancy below 50 years of age in 2007? Think about whether you are (filtering) rows or (selecting) columns. Replace [LESS THAN] with the logical operator for "less than" and [EQUAL TO] with the logical operator for "exactly equal to"

    ```{r part1-ex2, eval = FALSE}

    gapminder %>% 
        ___( lifeExp [LESS THAN] 50, year [EQUAL TO] ___ )

    ```


3. Which country had the higher life expectancy in 1977: Ireland or Brazil? Replace [OR] with the logical operator for "or" and [EQUAL TO] with the logical operator for "exactly equal to" in 3 places

    ```{r part1-ex3, eval = FALSE}

    gapminder %>% 
     filter( year [EQUAL TO] ___,
             country [EQUAL TO] "___" [OR] 
             country [EQUAL TO] "___" )

    ```


4. 

a. Sort the countries by life expectancy  to find the ones with the lowest life expectancy.  I have piped the result to the function `slice_head` from `dplyr` at the end of the pipeline to just print the first five rows.  

    ```{r part1-ex5, eval = FALSE}

    gapminder %>% 
       arrange(___) %>% 
                  slice_head(n = 5)  #only print top 5 rows             

    ```


b. Sort the countries in descending order of `gdpPercap` in 2007 and view the 5 countries with the highest values of GDP per capita.

    ```{r part1-ex5b, eval = FALSE}

    gapminder %>% 
       filter( ) %>% 
        arrange(___(gdpPercap)) %>% 
                       slice_head(n = 5)
        

    ```

5. Create a new variable called `gdp` which recovers each country's GDP by multiplying the population `pop` and GDP per capita `gdpPercap`. Add the new variable to the `gapminder` data frame. Examine the dataframe in your Enviromment. What do you notice about where the new variable is located? 

    ```{r part1-ex6, eval = FALSE}
gapminder <- gapminder %>% 
                         mutate( gdp = ___ * ___ )

    ```

**YOUR TURN: Answer the remaining problems in Part 1 by collaborating with your neighbors in small groups. Create a new code chunk below each question and write code to answer it. **


6. Which countries had a life expectancy (`lifeExp`)  greater than or equal to 80 in 2002?


7. What was the most populous country in Europe in 1992?  (You will want to first extract the rows from the data for Europe in 1992, then arrange in descending order on `pop`)


8. Which country in the Americas had the shortest life expectancy in the year 1962? 


9. Create a new variable `log_pop` which is the natural logarithm (`log`) of the population variable. Add the new variable to the gapminder data frame. (Then click on the `gapminder` dataframe in the Environment to see the column you just added!)


10. Which country had the higher GDP per capita (`gdpPercap`) in 1977? United States or United Kingdom? (Hint: you will want to extract the rows of data for these two countries in 1977)




* * *

## Part 2: count, distinct, group_by, summarise

We now return to the `gapminder` data frame to answer more questions using the data verbs we learned today: `count`, `distinct`, `group_by` and `summarise`. Be sure to load the packages **tidyverse** and **gapminder** in our interactive environment by submitting the code in the **setup** code chunk at the top of the document. 


1. How many unique countries does the data set contain? 
 


2. In the following code chunk, I have created a new variable called `LifeExp_cat` which categorizes `lifeExp` according to whether it is below 50 or at least 50. Print a frequency distribution of `lifeExp_cat` for the year 2002.


    ```{r part2-ex2, eval=FALSE}

    gapminder <- gapminder %>% 
               mutate( lifeExp_cat = ifelse(lifeExp < 50, 
                                            "below 50", 
                                            "at least 50") )

    ```

3. Summarize the life expectancy `lifeExp` by calculating the mean.

 
4. Summarize life expectancy `lifeExp` in Africa in 1952 by calculating the mean.

5. Summarize life expectancy `lifeExp` in 1952 for each continent separately by calculating the mean.



6. Summarize life expectancy `lifeExp` in 1952 for each continent separately by calculating the mean and median and sort the results in descending order by median life expectancy. 



6. Create a new variable called `popMil` which is the population in millions. For example, 8425333 is 8.425333 millions.  Use this new variable to calculate the total population in the world for each year. 



**Participation 2: Please do your calculations for the remaining problems in Part II below and then record your answers in CANVAS for the quiz titled Participation 2**




7. Summarise the life expectancy `lifeExp` in 1952 by calculating the mean.



8. Summarise the life expectancy `lifeExp` in Asia in 1952 by calculating the mean. 


9. Summarise the life expectancy `lifeExp` for each continent in 2002 by calculating the median. Sort your results in descending order.


10.  Create a new variable called `gdp` which recovers each country's GDP by multiplying the population `pop` and GDP per capita `gdpPercap`. Summarise the `gdp` for each country by calculating the total across all the years. Which 5 countries have the highest totals?


 

